---
title: "**terapadog**: Translational Efficiency Regulation Analysis & Pathway Analysis with Down-weighting of Overlapping Genes"
author:
- name: "Gionmattia Carancini"
  affiliation:
  - "CRT Genomics Data Sceince"
  - "LAPTILab, Uninversity College Cork, Ireland"
  email: gionmattia@gmail.com
date: "`r Sys.Date()`"
abstract: >
  An easy walktrough to learn how to use terapadog.
output: BiocStyle::html_document
vignette: >
  %\VignetteIndexEntry{terapadog: Translational Efficiency Regulation Analysis & Pathway Analysis with Down-weighting of Overlapping Genes}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

 

*Per Nonna Ezia, arrivederci e grazie di tutti i dolci.*

# Introduction

Welcome to the official (and only) documentation about TERAPADOG 
(***T**ranslational **E**fficiency **R**egulation **A**nalysis & **P**athway **A**nalysis with **D**own-weighting of **O**verlapping **G**enes*).

### What is TERAPADOG?

To put is short, TERAPADOG is an implementation of the classic PADOG aglorithm (Pathway Analysis with Down-weighting of Overlapping Genes), but on top of the results of a Differential Translational Analysis (executed with DeltaTE).
Now that the scientific jargon is out, let's delve a bit more into the topic.

Works on human only!

### Differential Translation-what?

In the last decade, experimental techniques such as Polysome profiling 
(Polysome-Seq) and Ribosome profiling (Ribo-Seq) have widely shown how gene
expression is not regulated only at the *transcriptional level* (through the 
increase/decrease of the pool of mRNA species in the cell), but also at the 
*translational level* (through the active increase/decrease of the *efficiency*
these mRNAs are translated to protein).

As a result of this, many packages have been developed to study gene regulation 
from this additional perspective, trying to characterize the changes in the
*Translational Efficiency* (TE) each mRNA could undergo between conditions, an
analysis which can be commonly referred to as 
*Differential Translation Analysis* (DTA).

The results of a DTA are usually a list of genes, each categorized into a
different *Regulatory Mode* depending on the changes in mRNA levels and TE
between the two conditions.
Such an output, although informative per se, but is also be quite complex to 
interpret when it comes to link together the different genes to infer the
overarching metabolic shifts happening in the experiment.

### The missing link

As stated above, a list of hundreds (if not thousands) of genes can be 
difficult to interpret from a systemic perspective.
A possible strategy to ease this step is to map this information on biological
pathways, so to obtain an overarching perspective showing how the different 
genes undergoing TE regulation are linked together.

It is from this idea that ***terapadog*** was conceived: to be able to perform a gene set analysis (applying the same logic as used in the PADOG package, see references for additional info) on the sets highlighted as being actively under Translational Efficiency changes.

### The bigger picture: integration within ReactomeGSA

While ***terapadog*** can be run as a tool on its own (as shown in this 
vignette), it was originally conceived to be an additional analysis method to 
the ReactomeGSA package, so to leverage on the visualisation methods and tools
offered by Reactome.

That is to say, while ***terapadog*** on its own allows the user to study genes 
that are undergoing changes in their TE and the pathways they are involved in, 
it is in its integration with the ReactomeGSA that it truly shines, thanks to 
the advanced framework and user-friendly visualization.

### Suggested Reading

For more information regarding PADOG or DeltaTE and Regulatory Modes, please 
refer to the paper mentioned in the references section of this vignette.

_About Differential Translation Analysis:_

Chothani S, Adami E, Ouyang JF, Viswanathan S, Hubner N, Cook SA, Schafer S, 
Rackham OJL. deltaTE: Detection of Translationally Regulated Genes by 
Integrative Analysis of Ribo-seq and RNA-seq Data. Curr Protoc Mol Biol. (2019). 

[https://pubmed.ncbi.nlm.nih.gov/31763789/]

_About PADOG:_

Tarca AL, Draghici S, Bhatti G. et al. Down-weighting overlapping genes improves 
gene set analysis. BMC Bioinformatics (2012). 

[https://doi.org/10.1186/1471-2105-13-136]


_About the ReactomeGSA:_

Griss J, Viteri G, Sidiropoulos K, Nguyen V, Fabregat A, Hermjakob H,
ReactomeGSA - Efficient Multi-Omics Comparative Pathway Analysis, Molecular 
& Cellular Proteomics (2020).

[https://doi.org/10.1074/mcp.TIR120.002155.]



# Installation

To install the terapadog package, execute the following code:

```{r install, echo = TRUE, eval=FALSE}
if (!require("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

BiocManager::install("terapadog")
```

# Package Structure



# Analysis Walktrough

This section will guide you thorough the various steps of the analysis, using
some testing data.
The data used at this stage is the same used by Chotani S. et al. in their paper
on DeltaTE and has been retrieved from their github repository.
Any manipulation on the data has been detailed in the *extdata_info.md* file 
(see scripts folder of the package).

[https://github.com/SGDDNB/translational_regulation.git]

## Input data: Formatting and Extension

First thing first, you need to make sure your input data is in the right format.

**In order to execute the analysis, it is mandatory input files are correctly 
formatted, so follow the guidelines detailed below.**

1) _**Samples Naming**_

In a DTA, you will have RNA-Seq counts and Ribo-Seq counts coming from the same
sample. 
That is to say, for "Sample A" you will have a set of gene counts obtained from
RNA-Seq and a set obtained from the Ribo-Seq protocol.

**It is mandatory for the sample name to be the same in both RNA-Seq and 
RIBO-Seq count files.**

*Example of RNA-Seq file:*

| GeneId | Sample_A | Sample_B | Sample_C | Sample_D |
| ---------- | ---------- | ---------- | ---------- | ---------- |
| ENSG0001 | 657 | 598 | 645 | 603 |
| ENSG0002 | 342 | 389 | 350 | 329 |

```{r, echo =FALSE}
# Tutorial Example
rna_counts <- system.file("extdata", "rna_counts.tsv", package = "terapadog")
rna_data <- read.csv(rna_counts, sep = "\t")

# Check the data
head(rna_data)
```

*Example of Ribo-Seq file:*

| GeneId | Sample_A | Sample_B | Sample_C | Sample_D |
| ---------- | ---------- | ---------- | ---------- | ---------- |
| ENSG0001 | 100 | 102 | 120 | 115 |
| ENSG0002 | 700 | 658 | 712 | 707 |


```{r, echo=FALSE}
# Tutorial Example
ribo_counts <- system.file("extdata", "ribo_counts.tsv", package = "terapadog")
ribo_data <- read.csv(ribo_counts, sep = "\t")

# Check the data
head(ribo_data)
```

Yes, they look almost identical in the formatting, with only the count values
as a striking difference.
It is intended, don't worry. 

2) _**Sample Info file**_
This file is basically a table where each Sample is linked to the experimental
condition you are assessing (for instance, control and disease).
It has only two columns:
- _SampleName_, the name of the sample.
- _Condition_, the experimental condition for said sample.

**It is mandatory the columns of this file are named with these specific names
and capitalization**

*Example of Sample info file:*

| SampleName | Condition | 
| ---------- | ---------- | 
| Sample_A | ctrl | 
| Sample_B | disease |
| Sample_C | ctrl |
| Sample_D | disease |

```{r echo=FALSE}
# Tutorial Example
sample_info_path <- system.file("extdata", "sample_info.tsv",
                                package = "terapadog")
sample_info <- read.csv(sample_info_path, sep = "\t")

# Check the data
sample_info
```
If the SampleName was different between RNA-Seq counts and Ribo-Seq counts, you
would have to compile 16 rows instead of just 8 in the sample info file. 
This is one of the (user-wise) reasons why the names must match.

3) _**File Format**_
Your count file must be in either a comma-separated values (.csv) or a 
tab-separated values (.tsv). Other extensions are currently not supported by
terapadog.

## prepareTerapadogData

Now that your input files are ready to go, we can start the terapadog workflow.
First thing first, you will have to call the function 
***preparedTerapadogData*** to load the files in your R environment and start
the preprocessing operations.

***preparedTerapadogData*** requires the following argument as _string_ inputs:

- _The path to the RNA-Seq counts file._
- _The path to the Ribo-Seq counts file._
- _The path to the sample info file._
- _The name of the condition you want to use as baseline in the comparison._

  For instance, if you have "control" and "disease" as conditions, "control" is 
  the baseline.
- _The name of the condition you want to use as target in the comparison._

  For instance, if you have "control" and "disease" as conditions, "disease" is 
  the target.

For tutorial purposes, this package comes with the three files you will need
to carry out a full terapadog analsysis.

- An RNA-Seq count file (*rna_counts.tsv*), which contains the mRNA-Seq derived 
counts for each gene in exam.
- A Ribo-Seq count file (*ribo_counts.tsv*), which contains the Ribo-Seq derived 
counts for each gene in exam.
- A sample info file (*sample_info.tsv*), which links each sample to a condition 
(in this case *1*  or *2*).



```{r}
# Load the library
library(terapadog)

# Read the paths in your R environments
path_rna_counts <- system.file("extdata", "rna_counts.tsv", package = "terapadog")
path_ribo_counts <- system.file("extdata", "ribo_counts.tsv", package = "terapadog")
sample_info_path <- system.file("extdata", "sample_info.tsv",
                                package = "terapadog")

# Call prepareTerapadogData
prep_data <- terapadog::prepareTerapadogData(path_rna_counts, path_ribo_counts,
                                  sample_info_path,"1", "2")

```

The result of this function will be a list with two elements:

- A matrix containing the counts from the RNA-Seq and Ribo-Seq files, formatted.
- A dataframe with all the info relative to the experimental design, which have 
been derived from sample_info, the two count files and the specified comparison.

```{r}
# Extract the matrix
expression.data <- prep_data$expression.data
print(head(expression.data))

# Extract the experimental design dataframe
exp_de <- prep_data$exp_de
print(exp_de)
```

## Executing the analysis

## Saving the results

## Something

# References

Testing data from

deltaTE tool

lfcShrink from apeglm

PADOG

ReactomeGSA

# SessionInfo

```{r session-info, echo = FALSE}
# Displays the session information
sessionInfo()
```
