---
title: "Translational Efficiency Regulation Analysis & Pathway Analysis with Down-weighting of Overlapping Genes (TERAPADOG)"
author: "Gionmattia Carancini"
date: "`r Sys.Date()`"
output: BiocStyle::html_document
vignette: >
  %\VignetteIndexEntry{terapadog: Translational Efficiency Regulation Analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# Introduction
Welcome to the official (and only) documentation about TERAPADOG 
(***T**ranslational **E**fficiency **R**egulation **A**nalysis & **P**athway 
**A**nalysis with **D**own-weighting of **O**verlapping **G**enes*).

### What is TERAPADOG?

To put is short, TERAPADOG is an implementation of the classic PADOG aglorithm
(Pathway Analysis with Down-weighting of Overlapping Genes), 
but on top of the results of a Differential Translational Analysis executed 
with DeltaTE.

### Differential Translation-what?

In the last decade, technologies like Polysome profiling (Polysome-Seq) and 
Ribosome profiling (Ribo-Seq) have shown that gene expression can be regulated 
at the translational level (through the modulation of Translational Efficiency,
TE), rather than only at the transcriptional level (through the 
increase/decrease of a specific mRNA's levels).

As a results of this, many packages have been developed to study gene regulation 
not only at the transcriptional level, but also translational. The results of 
these analyses are usually in the form of gene lists, where each gene is 
assigned a regulatory mode in regards of the changes in its mRNA level and 
Translational Efficiency (TE).

### The implementation with PADOG

Usually, results as a set of genes (especially if numerous) can be quite 
unintuitive to interpret. A possible strategy is to contextualize this 
information on biological pathways, so to have a more overarching perspective on
the processes that are being actively regulated through translation efficiency.

It is from this idea that TERAPADOG was conceived: to be able to perform a gene 
set analysis on the sets highlighted as being actively under Translational 
Efficiency changes. 

### The bigger picture: integration within ReactomeGSA

While TERAPADOG can be run as a tool on its own (as shown in this vignette), it 
is also conceived to be an additional analysis method to the ReactomeGSA package,
where the results of the analysis can be categorized into Regulatory Modes based
on their changes in TE and visualized on biological pathways thanks to Reactome's
tools and sources.

That is to say, TERAPADOG on its own allows the user to study genes that are 
undergoing changes in their TE.
In its integration with the ReactomeGSA things are pushed further, giving the 
user a more comprehensive picture that also encompasses the directionality of
the change (in mRNA or TE) and the possible regulatory modes arising from the
combination of the two.

### Additional Reading

For more information regarding PADOG or DeltaTE and regulatory modes, please
refer to the paper mentioned in the references section

# Installation

To install the terapadog package, execute the following code:

```{r install, echo = TRUE, eval=FALSE}
if (!require("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

BiocManager::install("TERAPADOG")
```
# Package Structure

# Walktrough 

## Preparing the input data

## Executing the analysis

## Saving the results

## Add script to get the gene FC and assign a RegMode?

# References

# SessionInfo

```{r session-info, echo = FALSE}
# Displays the session information
sessionInfo()
```
